---
title: "Astrocyte reactivity: RNA-seq data analysis"
output: 
    pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
library(tidyverse)
library(Biobase)
library(GEOquery)
library(data.table)
})
```


## Orignal paper title:
Modulation of astrocyte reactivity improves functional deficits in mouse models of Alzheimer’s disease
https://doi.org/10.1186/s40478-018-0606-1


## Expring GEO (Gene Expression Omnibus)
GEO - Gene Expression Omnibus \
GDS - GEO DataSet \
GSE - GEO Series \
GPL - GEO Platform \
https://www.ncbi.nlm.nih.gov/geo/info/faq.html \

Download GEO data and create the GEOquery object
```{r geo, echo=T, results=F, warning=FALSE, cache=T, message=F}
gse <- getGEO('GSE108520')

```
\  

Obtaining samples matadata:
```{r echo=T}

class(gse)
length(gse)

gse <- gse[[1]]
class(gse)

pheno     <- pData(gse) ## print the sample information

glimpse(pheno, width=80)
```

\  
\  

Now we take GSE object:
```{r geo2, cache=T, echo=T,  message=F}

geo_dat <- getGEO('GSE108520', destdir=".", GSEMatrix=F, AnnotGPL=T)

```

```{r}
mode(geo_dat)
class(geo_dat)
```
We are sure **geo_dat** is **GSE** Class
\  


### Exploring papaer information
```{r}
meta <- Meta(geo_dat)
attributes(meta)

```

\  

Summary:
```{r echo=T,  results=F, message=F, warning=F}
meta$summary
```
We analyzed the transcriptional profile of  astrocytes from:
1) WT mice infected with AAV-GFP
2) reactive astrocytes from 9-month old APP/PSdE9 mice infected with AAV-GFP
3) de-activated astrocytes from 9-month old APP/PSdE9 mice infected with AAV-SOCS3
"We show SOCS3 normalizes the inflammatory profile of APP astrocytes

\  

Experiment type:
```{r echo=T,   results=F, message=F, warning=F}
meta$type
```
Expression profiling by high throughput sequencing

\  

```{r echo=T,   results=F, message=F, warning=F}
meta$overall_design
```
Total RNA was extracted from GFP+ astrocytes isolated by FACS from WT and APP/PS1dE9 mice injected with an AAV targeting astrocytes and encoding GFP alone (controls, N=7 WT-GFP, N=4 APP-GFP) or SOCS3 and GFP (N=5 APP-SOCS3, same total viral load). Non GFP+ cells (including microglia, neurons, non infected astrocytes, called OTHER) were analyzed as  well, in 3 samples of the control WT-GFP group.


\newpage

## Data Analysis

\ 

Getting data:
```{r echo=T,   results=F, message=F, warning=F}
data_file <- meta$supplementary_file
dat <- read_delim(data_file, delim = "\t")
#dat <- read_delim("GSE108520_Deseq2_normalized_gene_expression_with_annotations.txt.gz", delim="\t")
```


```{r echo=T, results=T, cache=T}
glimpse(dat, width=80)

```

\  

Format data:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
names(dat)
edat_raw <- dat %>% select(-coordinates, -symbol)
edat_raw <- edat_raw %>% column_to_rownames(var='FastDB Stale ID')
edat_raw <- edat_raw[,sort(names(edat_raw))]

dim(edat_raw)
## gene names in rows
## samples in columns
edat_raw[1:5,1:4]

summary(edat_raw[,1:4])
```

\  

Preparing datastes, metadata with four groups:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
## we colud parse samples metadata just from samples name, but let's do it form GEO metadata
pheno %>% select(title, description) 

pdf4       <- pheno %>% select(title) 
pdf4$title <- pdf4$title %>%  str_replace_all("-","_")

pdf4$group <- str_split(pdf4$title, "_\\d", simplify=T)[,1] %>% 
  str_replace_all("Astro_","a") %>% 
  str_replace_all("Other_","o")

pdf4$group <- as.factor(pdf4$group)
names(pdf4) <- c("sname", "group")
rownames(pdf4) <- pdf4$sname
pdf4 <- arrange(pdf4, sname)

pdf4 %>% dplyr::count(group)

pdf4
```

\ 

Preparing metadata with three groups; we remove "other" types as they are not astrocytes, so thier experssion obvioulsy will be different:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
pdf3<- pdf4 %>% filter(!group=='oWT_GFP')
pdf3 <- droplevels(pdf3)
pdf3 %>% dplyr::count(group)

```

\  

Filtering expression counts and log transformation
```{r echo=T,   results=T, message=F, warning=F, cache=T}
## remove low expressed data
edat10  <- edat_raw[rowMeans(edat_raw) > 10, ] %>% arrange(rownames(.))
## wa want fold changes, but we can't do log2(0) so we add 1
edatlog4 <- log2(as.matrix(edat10) + 1) %>% as.data.frame()

## let's create only data set for three groups
edatraw3 <- edat_raw %>%  select(-Other_WT_GFP_1, -Other_WT_GFP_2, -Other_WT_GFP_3) %>% arrange(rownames(.)) 
edatraw3 <- edatraw3[rowMeans(edatraw3) > 10, ] 
edatlog3 <- log2(as.matrix(edatraw3) + 1) %>% as.data.frame()

summary(edatlog3[,1:4])

```

\  

Let's make sure the names are aligned
```{r echo=T,   results=T, message=F, warning=F, cache=T}
all.equal(colnames(edat_raw), pdf4$sname)
all.equal(colnames(edatlog4), pdf4$sname)
all.equal(colnames(edatraw3), pdf3$sname)
all.equal(colnames(edatlog3), pdf3$sname)

dim(edat_raw)
dim(edatlog4)
dim(edatraw3)
dim(edatlog3)

```
\  

Let's make some data in tidy form:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
etidy <- gather(edatlog4, key="sname", value="expr") %>% arrange(sname)
etidy <- left_join(etidy, pdf4)
```

\  

Let's see data summary:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
ggplot(etidy, aes(x=0,y=expr, fill=group)) + 
  geom_boxplot()  + 
  facet_grid(~sname) + 
  ylab("log2(counts)") +
  xlab("")
```

\  

Summary per group:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
ggplot(etidy, aes(x=0,y=expr, fill=group)) + 
  geom_boxplot()  + 
  facet_grid(~group) + 
  ylab("log2(counts)") +
  xlab("")
```

\

Histograms per group:
```{r echo=T,   results=T, message=F, warning=F, cache=T}

ggplot(etidy, aes(x=expr, fill=group)) + 
  geom_histogram(bins="50")  + 
  facet_grid(~group) 
```

\ 

Density plots:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
ggplot(etidy, aes(x=expr, colour=sname)) + 
  geom_density()  

ggplot(etidy, aes(x=expr, colour=group)) + 
  geom_density()  
```

\  

Heatmap:
```{r echo=T,   results=T, message=F, warning=F, cache=T}

library(pheatmap)

corMatrix <- cor(edatlog4)
pheatmap(corMatrix, annotation_col = select(pdf4, -sname))      

```


PCA:
```{r echo=T,   results=T, message=F, warning=F, cache=T}
library(ggplot2)
library(ggrepel)
library(stats)

expr_pca<-prcomp(t(edatlog4), center = TRUE, scale. = TRUE) 
summary(expr_pca)

screeplot(expr_pca, type = "l", npcs = 10, main = "Screeplot of the first 10 PCs")

imp <- summary(expr_pca)$importance
pc1 <- round(imp["Proportion of Variance","PC1"] *100, digits=0)
pc2 <- round(imp["Proportion of Variance","PC2"] *100, digits=0)

cbind(pdf4, expr_pca$x) %>% 
 ggplot(aes(x = PC1, y=PC2, col=group, label=rownames(pdf4) ) ) + 
 geom_point() + 
 geom_text_repel() +
 ylab(paste0("PC2(",pc2,"%)")) + 
 xlab(paste0("PC1(",pc1,"%)"))

```

We see, that other group has a bit different density profiles and cluster together. It is expected, as these are different cell types. We exclude them from differential expression analysis



\newpage

### Differential Expression for many groups

Many sources recommend to use linear models to find relations in RNA-seq count data, however in such scenario the data should be normally distributed (or at least the LM's residuals should). In the paper they use ANOVA or Kruskal-Wallis tests, depending on assumptions fulfillment.

\  

Here I try to use Generalized Linear Model, as RNA-Seq use to be not normal. 
Firstly let's test a normality of some random sample (if just one sample is not normally distributed, we can't use parametric testes or linear models)

```{r}
shapiro.test(sample(edatraw3$Astro_APP_GFP_2, 5000))

shapiro.test(sample(edatlog3$Astro_APP_GFP_2, 5000))

```
Shapiro-Wilk's p-value is less than 0.01 in both datasets (raw counts and log ratios), so should use non-parametric approaches

Let's check a qq plot:
```{r qq, cache=T}
library(ggpubr)

ggqqplot(data=edatlog3, x="Astro_APP_GFP_2") 
ggqqplot(data=edatraw3, x="Astro_APP_GFP_2") 

```
Both plots are concordant with the results of SW test.

\  

We can do the same test for some random genes (across all samples)
```{r qq2, cache=T}
test1 <- as.numeric(edatlog3['GSMG0032532',])
test2 <- as.numeric(edatlog3['GSMG0026079',])
shapiro.test(test1)
shapiro.test(test2)


```

\newpage

GLM:
Crate a description and contrast matrix: we are interested in differences in any group pairs
```{r}
suppressPackageStartupMessages({library(edgeR);library(limma)})

des_mat <- model.matrix(~ group + 0, data = pdf3)
colnames(des_mat) <- stringr::str_remove(colnames(des_mat), "group")

print(des_mat)


contrast_matrix <- makeContrasts(
  "aAPP_GFPvs" = aAPP_GFP  - aWT_GFP ,
  "aAPP_SOCSvs"= aAPP_SOCS - aWT_GFP ,
  "SOCS_GFAP"  = aAPP_SOCS - aAPP_GFP ,
  levels = des_mat
)

print(contrast_matrix)
```

\  

Let's be sure again that the names in data and matadata are aligned
```{r}
all.equal(colnames(edatraw3), pdf3$sname)

```

\newpage

For GLM we use raw counts data:
```{r glm, echo=T,   results=T, message=F, warning=F, cache=T}
# https://bioinformatics-core-shared-training.github.io/RNAseq-R/rna-seq-de.nb.html
# https://rpubs.com/bman/79395

# dge <- DGEList( counts=edatlog3, group=pdf3$group, lib.size=colSums( edatlog3 ) )
dge <- DGEList( counts=edatraw3, group=pdf3$group, lib.size=colSums( edatraw3 ) )
dge <- calcNormFactors( dge )

dge <- estimateGLMCommonDisp( dge, des_mat )
dge <- estimateGLMTrendedDisp( dge, des_mat )
dge <- estimateGLMTagwiseDisp( dge, des_mat )

fit <- glmFit( dge, des_mat )

glms <- glmLRT( fit, contrast=contrast_matrix )

diffs <-topTags( glms, n = nrow(glms)) %>% as.data.frame() %>% filter(FDR<0.01) %>% rownames_to_column("genes")


gene_map <- dat %>% select(`FastDB Stale ID`, symbol)
colnames(gene_map) <- c("genes", "symbol")

diffs <- select(diffs, genes, FDR)

left_join(diffs,gene_map) %>% select(genes, symbol, FDR)
  
```

\  

We discovered `r dim(diffs)[1]` significant results. The next step could be GO and KEGG anntations, functional analysis etc. but this is out of the scope of this project.
At the end let's visualize some randomly selected results (gene expressions)

```{r, message=F, warning=F}

plot_gene_expr <- function(gene_id) {

  top_gene_df <- edatraw3 %>%
    # Extract this gene from `expression_df`
    dplyr::filter(rownames(.) == gene_id) %>% as.matrix() %>% 
    # Transpose so the gene is a column
    t() %>%
    # Transpose made this a matrix, let's make it back into a data.frame like before
    data.frame() %>%
    # Store the sample ids as their own column instead of being row names
    tibble::rownames_to_column("sname") %>%
    # # Join on the selected columns from metadata
    dplyr::inner_join(dplyr::select(
      pdf3,
      sname,
      group
    ))
  
  ggplot(top_gene_df, aes_string(x = "group", y = gene_id, color = "group")) +
    geom_jitter(width = 0.1, height = 0, size=5) + 
    theme_classic() 
}


plot_gene_expr("GSMG0025455")
plot_gene_expr("GSMG0013278")
plot_gene_expr("GSMG0013663")
plot_gene_expr("GSMG0030665")




```









